---
title: "S20_802"
author: "Asela Sathsara"
date: "2025-07-11"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

In a supermarket in order to maximize its profit and to get run everyday shop life smoothly it is important to know how many customers arrive per day , how much time they spent shopping , how much time it would take them to get serviced by a cashier and most importantly if the available no of cashiers are enough to handle all the customers per day

In order to find all those information it is best to do a simulation using the obtained data from the supermarket

This report presents a discrete-event simulation model of a supermarket checkout system, simulating one business day (13 hours) with four parallel checkout counters. Customers arrive according to a Poisson process, engage in shopping with log-normal distributed duration, and join the shortest queue upon completing shopping. Cashiers have different service efficiency levels, modeled with exponential distributions. By using this simulation we aim to measure factors like how many customers arrive each day , the average waiting time they have to wait before leave the shop and many more.


# Methodology


## variables 

Let's first identify and define the variables that full describe the status of the system at any point of time

1. Number of customer arrivals by time t
2. Number of customer departures by time t
3. Number of customers in the system to be served by a cashier
4. Number of idle cashiers in time t
5. Arrival time of the next customer
6. Departure time of the customer currently served by each cashier
   (since there are 4 cashiers with different service times we need to consider each cashier separately)
 
 
## Events occur during the simulation

Now lets define all possible events that can occur during this simulation that can affect into the overall performance of simulation

1. Customer arrival

   \t-> A new customer arrive at the supermarket
   
   \t-> shopping time is generated using log normal
   
   \t-> A shopping end event is scheduled
   
2. End of Shopping

   \t-> customer finishes the shopping
   
   \t -> find the shortest queue and join the queue
   
3. Queue Selection

   \t -> customer find a queue with shortest length and join the queue
   
4. Service start

   \t -> If the queue is empty immediately the service start otherwise he need to wait until his chance is there
   
   \t -> the service time is generated by exponential distribution based on cashier

5. Service end

   \t -> Cashier finishes servicing the customer.

   \t -> If there’s another customer in the queue, they begin service immediately (new Service Start).

   \t -> If no one is in queue, cashier becomes idle.


## The data given for the simulation

open time = 8.00 a.m.

close time = 9.00 p.m.

total time shop opened for = 13* 60 = 780 minutes

customer arrival rate per minute = 5
 
shopping mean = 0 ( Log-normal distribution ) 

shopping standard deviation = 0.5 ( Log-normal distribution )

service rates per cashier = c(1/2.5, 1/3, 1/3.5, 1/4)


  
## Step by step how the simulation happen

1. Generate a customer arrival using exponential inter arrival times
and make sure to remove customers who arrive after 9.00 p.m. (i.e. 780 minute) . Using a while loop is recommended for that

2. Generate shopping time for each customer using log normal distribution .

3. Since data for each customer is needed make a data frame with necessary variables (columns) to  store them.

4. Initialize queues and servers

5. For each customer
 
  \t -> find the shortest queue and break ties randomly

  \t -> determine the service start time ( service can start immediately if the cashier is free )

  \t -> record the data into the data frame

6. Calculate the key parameters needed for the conclusion



```{r echo=FALSE, message=FALSE, warning=FALSE}
library(dplyr)

set.seed(1111)

#given information

open_time <- 0
close_time <- 13*60 # since there are 13 hours (13*60 = 780)

lambda <- 5  # arrival rate per minute

shopping_mean <- 0
shopping_sigma <- 0.5

cashier_service_rates = c(1/2.5,1/3,1/3.5,1/4) #service rates for cashiers


# let's generate arrivals

arrival_times <- c()
time <- 0

while (time <= close_time){
  interarrival <- rexp(1 , rate = lambda) # generating a exponential random number
  time <- time + interarrival
  
  if(time <= close_time){
    arrival_times <- c(arrival_times , time)
  }
  
}

no_of_customers <- length(arrival_times)


# shopping times

shopping_times <- rlnorm (no_of_customers , meanlog = shopping_mean , sdlog = shopping_sigma)
shopping_end_times <- arrival_times + shopping_times

# lets create a customer data frame in order to collect data for each customer

customers <- data.frame(
  id = 1:no_of_customers,
  arrival_time = arrival_times,
  shopping_time = shopping_times,
  shopping_end_time = shopping_end_times,
  counter = NA,
  queue_entry_time = NA,
  service_start_time = NA,
  service_end_time = NA
)

# Initializing queues and servers

queues <- list(rep(0, 0), rep(0, 0), rep(0, 0), rep(0, 0)) #in here queues are assigned for each vacant server
cashier_available_time <- rep(0, 4)


# simulating the event

for(i in 1:no_of_customers){
  t <- customers$shopping_end_time[i]
  
  
  # in here i decided to find the shortest queue so that customer can go to it
  # also ties are break randomly
  
  queue_lengths <- sapply(queues, length)
  minimum_queue <- which(queue_lengths == min(queue_lengths))
  
  selected <- sample(minimum_queue,1) # in order to break ties
  
  customers$counter[i] <- selected
  customers$queue_entry_time[i] <- t
  
  # if cashier is free we can proceed 
  
  if(cashier_available_time[selected] <= t){
    customers$service_start_time[i] <- t
  }else {
    customers$service_start_time[i]<- cashier_available_time[selected]
  }
  
  service_time <- rexp (1 , rate = cashier_service_rates[selected])
  customers$service_end_time[i] <- customers$service_start_time[i] + service_time
  cashier_available_time[selected] <- customers$service_end_time[i]  
  
}


#lets add 2 new colomns
customers <- customers %>%
  mutate(
    waiting_time = service_start_time - shopping_end_time,
    total_time = service_end_time - arrival_time
    
  )


#average waiting time per counters
avg_waiting <- customers %>%
  group_by(counter) %>%
  summarise(avg_waiting_time = mean(waiting_time))

# average total time in supermarket

avg_total_time <- mean(customers$total_time)


# utilizing the cashiers

cashier_utilisation <- sapply(1:4, function(i) {
  sum(customers$service_end_time[customers$counter == i] - 
        customers$service_start_time[customers$counter == i]) / 780
})

# total customers

total_customers <- nrow(customers)

```

```{r echo=FALSE}
library(ggplot2)
```





# Results and Discussion 

## required results 

#### * The average waiting time experienced by customers at each checkout counter


```{r echo=FALSE}
print(avg_waiting)
```


#### * The average time a customer spends in the supermarket

```{r echo=FALSE}
cat("\nAverage time spent in supermarket:", avg_total_time, "minutes\n")

```

#### * The utilisation of each cashier per day

```{r echo=FALSE}
print(data.frame(Cashier = 1:4, Utilisation = cashier_utilisation))

```


#### * The total number of customers served per day 

```{r echo=FALSE}

cat("\nTotal number of customers served:", total_customers, "\n")

```


## additional data obtained from the simulation 

First lets get the mean service time for each cashier because it would help us to get an idea about how each cashier operates

```{r echo=FALSE}

library(knitr)

mean_service_times <- 1 / cashier_service_rates


cashier_service_table <- data.frame(
  Cashier = paste("Cashier", 1:4),
  Service_Rates = round(cashier_service_rates, 4),
  Mean_Service_Time = round(mean_service_times, 2)
)

# Print the table
print(cashier_service_table)
```







```{r echo=FALSE}
ggplot(avg_waiting, aes(x = factor(counter), y = avg_waiting_time)) +
  geom_bar(stat = "identity", fill = "skyblue") +
  labs(title = "Average Waiting Time per Cashier", x = "Cashier ID", y = "Avg 
Waiting Time (min)")
```

In here it clearly show that the average waiting time for cashier 4 is the highest among them . it would be the result of having the minimum service rates and that will cause the mean service time for become high for cashier no 4







```{r echo=FALSE}
ggplot(customers, aes(x = shopping_time)) +
  geom_histogram(binwidth = 0.05,fill = "lightgreen", color = "black") +
  labs(title = "Distribution of Shopping Times", x = "Shopping Time (min)", y = 
"Count") 
```

As we can see most of the shopping times are between 0 to 1 minutes . In my theory it would another reason for higher queue sizes for each cashier . Because according to the table of mean service time of each cashier it would take at least minimum of 2.5 minutes mean time to give service in the cashier . But if the most of shopping time is less than 1 or 2 minutes it would certently make a big queue and result in to a higher waiting time








```{r echo=FALSE}
ggplot(customers, aes(x = waiting_time, fill = factor(counter))) +
  geom_density(alpha = 0.4) +
  labs(title = "Waiting Time Distribution by Cashier", x = "Waiting Time (min)", 
fill = "Cashier ID")
```

In here as we can see cashier 1 (red) has the highest density peak near the lower waiting times indicating that most customer served by the cashier 1 wait very little time . 

Cashier 2(Yellow) has a moderate peak and the curve extends further right than cashier 1. This shows that while many customers had short waits a noticeable number had longer waiting times.


Cashier 3 (Cyan) has a curve with broader and more spread out across a wide range of waiting times. This suggests a larger variation in how long customers waited.

Cashier 4 (Purple) Has a much wider and right-shifted distribution.The peak is relatively low, but the tail stretches far beyond 3000 minutes indicating cashier 4 is the slowest, causing long queues and significantly higher waiting times for many customers.




# Conclution

From the results obtained from the simulation run we can see that average time spent in the supermarket is equal to 1224.157 minutes and its nearly equal to 20 hours . This is not a acceptable for a supermarket since it can only operate for 13 hours per day

There fore this supermarket isn't at its optimum service capability

One thing that can change and try again is how many service cashiers are in the supermarket . If the no cashiers increase then there will be less queues in the supermarket . The thing is finding how much that increase should be for the no of cashiers . 

Other option is rejecting customer arrivals long before 9.00 p.m. it will make less profit but there will be less no of customers waiting to be serviced by cashiers now and that will lead to a less average waiting time for a customer .

In summary running the supermarket with these parameters will lead to a unacceptable waiting time for customers to get serviced and in order to get the service to be optimum some changes should happen .



# References , Softwares and packages used

```{r echo=FALSE}
citation()
citation("dplyr")
citation("ggplot2")
citation("knitr")
```

